\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage{xcolor}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  urlcolor=blue
}

\setstretch{1.15}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.75em}

\title{Occupation Level AI Exposure Analysis}
\author{Final Project}
\date{\today}

\begin{document}

\maketitle

\section*{Introduction}

This project was designed to study how artificial intelligence exposure appears across occupations and how that exposure relates to salary, automation risk, outlook, and other job characteristics. The work is based on occupation level data from Anthropic's Economic Index repository together with wage and occupation structure files. The final result is presented as a Streamlit dashboard with seven tabs, each of which serves a different analytical purpose.

The report follows that same dashboard structure. Instead of presenting the work as disconnected code or isolated charts, it explains what each tab is doing, what the main outputs mean, and why those outputs matter for the larger project question. The central idea throughout the project is simple: some occupations show a measurable AI exposure signal, and that signal can be compared with labor market indicators in a structured way.

\section*{Dataset and Preparation}

The dataset used in the project was built by merging three CSV files. The first file, \texttt{job\_exposure.csv}, contains Anthropic's occupation level \texttt{observed\_exposure} score. The second file, \texttt{wage\_data.csv}, contains salary, forecast, automation chance, job family, job zone, and occupation labels. The third file, \texttt{SOC\_Structure.csv}, provides major occupation group names from the standard occupation hierarchy.

The merge was done at the occupation level. Occupation codes from the wage file were standardized by removing decimal suffixes, and only the canonical rows ending in \texttt{.00} were kept so that the final dataset would contain one row per occupation rather than repeated sub specialty rows. After the merge and cleaning process, the final dataset contained 672 occupations and 25 columns.

Several cleaning steps were applied before analysis. Salary values smaller than 100 were treated as hourly wages and converted into annualized salary. Placeholder values of \texttt{-1} in the automation chance and job zone columns were treated as unknown rather than valid measurements. Readable labels were then created for job zone, bright outlook, green status, exposure group, and exposure intensity. As a result, the final dataset is easier to explain to someone who has never seen the raw files.

\section*{Tab 1: Summary}

The Summary tab is the entry point to the dashboard. Its purpose is to show that the dataset is valid, understandable, and ready for analysis. This tab is not intended to make a statistical argument by itself. Instead, it establishes the basic structure of the data and gives the user confidence that the later analysis is grounded in a clean occupation level table.

\subsection*{Data Validation and Summary}

The first element in this tab is the data quality table. This table reports the data type of every column, the number of missing values, the percentage of missing values, and the number of distinct values. This gives a quick view of which fields are numeric, which fields are categorical, and where missing data exists. It is especially useful for showing that columns such as automation chance and job zone had to be cleaned before being used in the analysis.

Next to the data quality table is the numeric summary table. This table reports count, mean, median, mode, standard deviation, minimum, quartiles, and maximum for each numeric variable. These values serve as the preliminary descriptive statistics required in the project instructions. In practical terms, this table is where the user sees the scale of the salary column, the spread of automation chance, and the range of observed exposure values.

\subsection*{Exposure Group Counts}

The first chart in the Summary tab shows how many occupations fall into the two exposure groups. One group contains occupations with \texttt{observed\_exposure} equal to zero, and the other contains occupations with \texttt{observed\_exposure} greater than zero. This chart gives an immediate sense of class balance and shows whether the exposure analysis is dominated by one group or whether both groups have enough occupations for comparison.

\subsection*{Top 10 Job Families by Average Exposure}

The second chart in the Summary tab ranks job families by their average observed exposure. This is a useful bridge between summary and exploration. It does not yet drill into individual occupations, but it already shows where AI exposure appears to be concentrated. For a reader or audience member, this chart helps establish the overall pattern before moving into the more detailed visual analysis in the next tab.

\section*{Tab 2: EDA}

The EDA tab is where the main exploratory work happens. The purpose of this tab is to uncover patterns rather than test formal hypotheses. Every visual here is trying to answer a version of the same question: which occupations appear more exposed to AI, and what other occupation characteristics move with that exposure?

\subsection*{Average Observed Exposure by Job Family}

This chart compares job families using their mean \texttt{observed\_exposure}. It is useful because it moves beyond isolated occupations and looks at broader categories of work. In this project, this visual helps show that exposure is more common in information rich and professional job families than in many manual or routine families.

\subsection*{Top Major Occupation Groups by Exposure}

This chart performs a similar comparison at the major group level using the occupation hierarchy from the SOC file. It gives a broader labor market view than the job family chart. The point of this visual is to show whether the exposure pattern still holds when occupations are grouped at a higher level. It also helps confirm that the findings are not being driven only by narrow labels in the wage file.

\subsection*{Salary vs Observed AI Exposure}

This scatter plot places salary on the horizontal axis and observed exposure on the vertical axis. Each point represents one occupation, and bubble size reflects job forecast. The purpose of the chart is to show whether more exposed occupations tend to sit in particular salary ranges and whether strong labor demand appears together with exposure. This is one of the most useful plots in the app because it brings together exposure, salary, and future demand in a single view.

\subsection*{Top Occupations by Automation Chance, Colored by AI Exposure}

This chart focuses directly on occupations with the highest automation chance. The bar length represents automation risk, while the color shows observed AI exposure. The purpose of this chart is to separate two ideas that are easy to confuse. A job can have high automation risk without having the highest observed AI exposure, and a job can have positive exposure without being among the most automatable. This visual supports one of the main ideas of the project: AI exposure and automation chance are related, but they are not identical.

\subsection*{Salary by Exposure Group}

This visual compares salary distributions between occupations with positive exposure and occupations with no exposure. It is a distribution level chart rather than a single number comparison, which makes it useful during exploratory analysis. The visual gives an early indication that the exposed group tends to have a higher salary distribution, which then motivates the formal salary hypothesis test in the next tab.

\subsection*{Average Exposure by Job Family and Job Zone}

This heatmap combines job family and job zone. The darker the cell, the higher the average observed exposure for that combination. This visual is useful because it adds the preparation dimension to the story. It allows the user to see that exposure does not vary only across industries or job families. It also varies with the level of preparation, education, training, and experience usually needed for the occupation.

\subsection*{Top Exposed and Low Exposure Occupations}

The final part of the EDA tab shows ranked occupation tables. One table lists the most exposed occupations, while the other shows low or zero exposure examples. These tables make the analysis concrete. Instead of talking only about averages, the app shows the actual occupation names that sit at the high and low ends of the exposure spectrum.

\section*{Tab 3: Hypothesis Test}

The Hypothesis Test tab moves from description to inference. The goal of this tab is to test whether the differences suggested by the exploratory visuals are statistically meaningful. The app currently includes two separate two sample t tests, each comparing occupations with positive exposure against occupations with no exposure.

\subsection*{Hypothesis 1: Salary and AI Exposure}

The first hypothesis asks whether the mean annualized salary differs between the two exposure groups. The null hypothesis states that the mean salary is the same for both groups. The alternative hypothesis states that the mean salary is different.

The tab reports the mean of each group, the sample size of each group, the t statistic, the p value, and the effect size. It also shows normality checks using Shapiro p values and a variance check using Levene's test. The accompanying violin plot makes it easy to see the salary distribution for both groups while the statistical output provides the formal conclusion.

This part of the tab is important because the EDA suggested a salary gap between exposed and non exposed occupations. The hypothesis test tells us whether that gap is large enough to be considered statistically meaningful in the sample.

\subsection*{Hypothesis 2: Automation Chance and AI Exposure}

The second hypothesis asks whether the mean automation chance differs between the same two exposure groups. The null hypothesis states that the mean automation chance is the same for exposed and non exposed occupations. The alternative hypothesis states that the mean automation chance is different.

This second test strengthens the project because it directly addresses a central conceptual question. The project is not only about whether exposed occupations earn different salaries. It is also about whether AI exposure and automation risk are telling the same story. By running a second t test on automation chance, the app turns that idea into a formal statistical comparison rather than leaving it as an informal observation from the EDA.

\section*{Tab 4: ML Exposure Classification}

The first machine learning tab uses supervised learning to classify occupations into two groups: no exposure and positive exposure. The model used here is a random forest classifier. The target variable is binary and is created from the observed exposure column. If observed exposure is greater than zero, the occupation is assigned to the positive exposure class. Otherwise, it remains in the no exposure class.

\subsection*{Classification Metrics}

This part of the tab reports accuracy, precision, recall, and F1 score. Each metric shows a different aspect of model performance. Accuracy gives the overall share of correct predictions. Precision shows how often predicted positive exposure occupations were actually positive. Recall shows how many of the true positive exposure occupations were found by the model. F1 score balances precision and recall.

The tab also reports the number of training rows and test rows. This makes the train test split visible and helps the user understand that the model was evaluated on held out data rather than on the same data used for fitting.

\subsection*{Confusion Matrix}

The confusion matrix shows the number of correct and incorrect predictions broken down by class. This is a useful complement to the summary metrics because it reveals whether the model is biased toward one class or whether it performs differently on exposed and non exposed occupations.

\subsection*{Feature Importance}

The feature importance chart identifies which variables were most useful in predicting exposure status. This is valuable because the classifier is not only being used for prediction. It is also being used as an exploratory machine learning tool to highlight which occupation characteristics are most associated with the presence of AI exposure in the data.

\section*{Tab 5: ML Salary Regression}

The second machine learning tab uses linear regression to predict annualized salary. The purpose of this tab is different from the classification tab. Here the model is not trying to predict exposure status. Instead, it is trying to model salary while including observed exposure as one of the predictors.

\subsection*{Regression Performance}

The first outputs in this tab are \(R^2\), mean absolute error, and root mean squared error. These metrics summarize how well the regression model predicts salary. They do not prove causal relationships, but they do show how well occupation features jointly explain salary variation in the data.

\subsection*{Largest Coefficients with P Values}

This part of the tab reports the largest coefficients in the model together with p values. The coefficients show the direction and relative size of the association between each feature and salary, holding the other included variables constant. The p values help identify which coefficients are statistically significant in the fitted linear model.

This is especially important for \texttt{observed\_exposure}. The project is interested in whether AI exposure is associated with salary after controlling for other job characteristics. The coefficient for observed exposure and its p value provide a more careful answer than a simple correlation or scatter plot alone.

\subsection*{Interpretation of the Exposure Coefficient}

The final part of the regression tab gives a plain language interpretation of the observed exposure coefficient. The wording is careful by design. The app describes the relationship as an association while holding the other variables constant. It does not claim that AI exposure causes salary changes. This distinction matters because the project is observational, not experimental.

\section*{Tab 6: Career Insight}

The Career Insight tab is a practical extension of the main analysis. Instead of only showing global patterns, it lets the user select a single occupation and view a data driven profile of that occupation. This tab is not meant to replace the statistical analysis. Its purpose is to translate the dataset into a more intuitive form for someone who wants to understand how a specific occupation looks within the broader labor market and AI exposure context.

\subsection*{Occupation Profile}

The tab displays the selected occupation's profile label, observed exposure, automation chance, annualized salary, job family, job zone, bright outlook status, and green status. These fields provide a compact view of the occupation using the same variables analyzed elsewhere in the app.

\subsection*{Relative Scorecard}

The central chart in this tab is a scorecard built on percentile based measures. The app converts observed exposure, automation chance, salary strength, and outlook strength into relative scores on a zero to one hundred scale. It then combines those into two summary ideas: an AI disruption score and a career opportunity score.

The scoring logic is intentionally simple and interpretable. AI disruption blends exposure and automation chance. Career opportunity blends salary, forecast, and bright outlook. The purpose is not to claim a universal career truth. It is to give the user a structured, reproducible summary of how a selected occupation sits inside this dataset.

\subsection*{Narrative Interpretation}

The Career Insight tab also generates a short textual interpretation of the selected occupation. This narrative is rule based rather than black box. It describes whether the occupation sits in the high, middle, or low range of exposure, automation chance, and opportunity indicators. In presentation terms, this is one of the most audience friendly parts of the dashboard because it turns the underlying analysis into an occupation specific takeaway.

\section*{Tab 7: Definitions}

The final tab exists for clarity. A dashboard can easily become confusing when it includes terms such as observed exposure, job zone, bright outlook, and exposure intensity. The Definitions tab makes the project more accessible to anyone who was not involved in building it.

\subsection*{Project Terminology}

This table explains the meaning of the key labels and variables used across the app. It includes terms such as \texttt{observed\_exposure}, positive exposure, no exposure, automation chance, job forecast, bright outlook, green label, job zone, exposure intensity, annualized salary, major occupation group, AI disruption score, and career opportunity score.

\subsection*{Job Zone Guide}

The tab also includes a short guide explaining the job zone labels from Zone 1 through Zone 5, along with the meaning of the unknown category. This is useful because job zone is one of the most important variables in the project, but it is not obvious to a new reader without explanation.

\section*{Conclusion}

The dashboard as a whole is designed to move from understanding the data to interpreting the broader meaning of the results. The Summary tab validates the structure of the data. The EDA tab reveals patterns across occupations, salary, preparation level, and automation chance. The Hypothesis tab tests whether the salary and automation differences between exposure groups are statistically meaningful. The two machine learning tabs then extend the analysis by predicting exposure status and salary. Finally, the Career Insight and Definitions tabs make the project more interpretable and more useful to a general audience.

Taken together, the project supports a more careful view of AI in the labor market. Occupations with measurable AI exposure are not simply the same occupations with the highest automation risk. In many cases, exposure appears in jobs that are information rich, preparation heavy, and relatively well paid. That is why the project treats exposure and automation as related but distinct concepts. The app is designed to make that distinction visible, testable, and easy to explain.

\section*{References}

Anthropic Economic Index. \url{https://huggingface.co/datasets/Anthropic/EconomicIndex}

Anthropic labor market impacts report. \url{https://www.anthropic.com/research/labor-market-impacts}

O*NET Online. \url{https://www.onetonline.org/}

\end{document}
